---
title: "Repriducible_Research_Assignment_1"
author: "Edoardo Pennesi"
date: "10 February 2019"
output:
  pdf_document: default
  html_document: default
---

### setting default options to include the codes

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading the necessarely libraries

```{r}
library(ggplot2)
library(plyr)
```



## 1. Loading and preprocessing the data

```{r}
# download the data if necessarely
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip" 
if(!file.exists("repdata_data_activity.zip")) 
{download.file(url, destfile = "repdata_data_activity.zip", method = "curl")}

# Unzip archive
unzip("repdata_data_activity.zip")

# Read the data into a data frame.
activity <- read.csv("activity.csv")

# use as.Date( ) to convert strings to dates 
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")

# check the structure of the data
str(activity)
```

## 2. What is mean total number of steps taken per day?

### 2.1 Calculate the total number of steps taken per day

```{r}

## summarizing total steps per date
tot.steps.day <- aggregate(steps ~ date, rm.na = TRUE, data = activity, FUN = sum)
colnames(tot.steps.day)<- c("Date", "Steps")
```



### 2.2 Plot histogram with frequency of steps by day 
```{r}
hist(tot.steps.day$Steps, breaks=20, xlab = "Steps", main = "Total Steps per Day")
```

### 2.3 Calculate and report the mean and median of the total number of steps taken per day

```{r}
## Mean of Steps
as.integer(mean(tot.steps.day$Steps))

## Median of Steps
as.integer(median(tot.steps.day$Steps))

```

The average number of steps taken each day was 10766 steps.

The median number of steps taken each day was 10765 steps.

## 3 What is the average daily activity pattern?

###3.1. Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
plot(aggregate(steps ~ interval, data = activity, FUN = mean), type = "l")
```

###3.2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max(activity$steps, na.rm = TRUE)
```


## 4. Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

## 4.1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
sum(is.na(activity))
```

- The total number of rows with steps = ‘NA’ is 2304.

### 4.2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r}
# calculate mean of steps per interval
avg.steps.int <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
# cut the 'activity' dataset in 2 parts (with and without NAs)
activity_NAs <- activity[is.na(activity$steps),]
activity_non_NAs <- activity[!is.na(activity$steps),]

#replace missing values in activity_NAs
activity_NAs$steps <- as.factor(activity_NAs$interval)
levels(activity_NAs$steps) <- avg.steps.int

#change the vector back as integer 
levels(activity_NAs$steps) <- round(as.numeric(levels(activity_NAs$steps)))
activity_NAs$steps <- as.integer(as.vector(activity_NAs$steps))

#merge/rbind the two datasets together
imputed_activity <- rbind(activity_NAs, activity_non_NAs)
```


### 4.4 Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
#Plotting parameters to place previous histogram and new one next to each other
par(mfrow = c(1,2))
#Plot again the histogram from the first part of the assignment
activity_steps_day <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
hist(activity_steps_day$steps, xlab = "Steps per Day", main = "NAs REMOVED - Total steps/day", col = "gray")
#Plot new histogram, with imputed missing values
imp_activity_steps_day <- aggregate(steps ~ date, data = imputed_activity, FUN = sum, na.rm = TRUE)
hist(imp_activity_steps_day$steps, xlab = "Steps per Day", main = "NAs IMPUTED - Total steps/day", col = "black")

imp_mean_steps <- mean(imp_activity_steps_day$steps)
imp_median_steps <- median(imp_activity_steps_day$steps)
```



### 4.5 Do these values differ from the estimates from the first part of the assignment? 

The mean of the imputed data is 10821 steps. 
The mean of the NON-imputed data is 10766 steps. 
The mean is 55 units higher when na values are repalced.

The new median of the imputed data is 11015 steps. 
The median of the NON-imputted data is 10765 steps. 
The median is 250 units higher when na values are repalced.


### 4.6 What is the impact of imputing missing data on the estimates of the total daily number of steps?

the overall shape of the distribution has not changed significantly.

## 5 Are there differences in activity patterns between weekdays and weekends?

### 5.1 Create a new factor variable

```{r}

# categorize Saturday and Sunday as factor level "weekend" and all the other days  as "weekday"
imputed_activity$dayType <- ifelse(weekdays(as.Date(imputed_activity$date)) == "Saturday" | weekdays(as.Date(imputed_activity$date)) == "Sunday", "weekend", "weekday")

#transform dayType variable into factor
imputed_activity$dayType <- factor(imputed_activity$dayType)


#Aggregate a table showing mean steps for all intervals, acrlss week days and weekend days
steps_interval_dayType <- aggregate(steps ~ interval + dayType, data = imputed_activity, FUN = mean)
#verify new dataframe 
head(steps_interval_dayType)


#add descriptive variables
names(steps_interval_dayType) <- c("interval", "day_type", "mean_steps")
```

### Make a panel plot containing a time series plot (i.e. type = "l"\color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}

#plot with ggplot2
plot <- ggplot(steps_interval_dayType, aes(interval, mean_steps))
plot + geom_line(color = "red") + facet_grid(day_type~.) + labs(x = "Intervals", y = "Average Steps", title = "Activity Patterns")
```

The plots show that the agerage steps taken by the subjects seems higher during the weekend with a peack between 8 am and 9 am in both cases
